---
title: "Making sense of investing for retirement"
date: "2020-05-15"
summary:
image:
  caption: Image by [NeONBRAND](https://unsplash.com/photos/JW6r_0CPYec)
  focal_point: 'center'
output:
  blogdown::html_page:
    toc: false
    toc_depth: 1
url_code: "https://github.com/iecastro/blogdown-site/blob/master/content/post/2020-05-15-ira-portfolio/index.Rmd"
links:
tags: [R, data viz]
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)

```


```{r libs}
library(tidyverse)
library(tidyquant)
library(patchwork)

```

These are my value allocations based on share closing prices on 5/13/2020, which will be used as weights for the portfolio analysis.

```{r}

weights <- tribble(
  ~acct, ~symbol, ~prop,
  "combined", "BND", 0.090,
  "combined", "DLR", 0.186,
  "combined", "EPR", 0.216,
  "combined", "VNQ", 0.202,
  "combined", "VTI", 0.119,
  "combined", "VTR", 0.141,
  "combined", "VXUS", 0.046,
  "ira", "BND", 0.140,
  "ira", "DLR", 0.289,
  "ira", "VNQ", 0.314,
  "ira", "VTI", 0.185,
  "ira", "VXUS", 0.072,
  "ind", "EPR", 0.605,
  "ind", "VTR", 0.395
)

```


```{r returns}
# set a start date for returns
start_date <- "2019-01-01"

# monthly returns for individual securities
returns_combined <- tq_get(unique(weights$symbol),  
                           get = "stock.prices",
                           from = start_date) %>% 
  group_by(symbol) %>%
  tq_transmute(adjusted, periodReturn, 
               period = "monthly",
               col_rename = "Ra")

# portfolio returns
get_returns <- function(return_df, weight_df, growth = FALSE){
  
  df <- return_df
  
  x <- weight_df
  
  # filter weights for correct acct
  wt <- filter(weights, 
               acct == x) %>% 
    select(-acct)
  
  tq_portfolio(df,
               assets_col = symbol,
               returns_col = Ra,
               weights = wt,
               col_rename = "returns",
               wealth.index = growth)
}
  

port_combined <- returns_combined %>% 
  get_returns("combined")

port_ira <- returns_combined %>% 
  filter(!symbol %in%
           c("EPR", "VTR")) %>% 
  get_returns("ira")

port_ind <- returns_combined %>% 
  filter(symbol %in%
           c("EPR", "VTR")) %>% 
  get_returns("ind")

# track growth
growth_ira <- returns_combined %>% 
  filter(!symbol %in%
           c("EPR", "VTR")) %>% 
  get_returns("ira", growth = TRUE)

growth_ind <- returns_combined %>% 
  filter(symbol %in%
           c("EPR", "VTR")) %>% 
  get_returns("ind", growth = TRUE)

```


```{r plot-01}
source("plot-functions.R")

p_combine <- plot_returns(returns_combined, port_combined) +
  labs(subtitle = "Combined portfolios")

p_combine

```


```{r plot-portfolios, fig.height=11, fig.width=12}
# returns plots
p_ira <- returns_combined %>% 
  filter(!symbol %in%
           c("EPR", "VTR")) %>% 
  plot_returns(port_ira) +
  labs(subtitle = "Roth IRA acct.")

p_ind <- returns_combined %>% 
  filter(symbol %in%
           c("EPR", "VTR")) %>% 
  plot_returns(port_ind) +
  labs(subtitle = "Indv. taxable acct.")

# growth plots
grow_ira <- plot_growth(growth_ira) +
  labs(subtitle = "Roth IRA acct. growth of $1")

grow_ind <- plot_growth(growth_ind) +
  labs(subtitle = "Indv. taxable acct. growth of $1")

# patchwork
p_combine / (p_ira +  p_ind) / (grow_ira + grow_ind)


```


```{r plot-10yr-portfolios, fig.height=11, fig.width=12}

start_date <- "2010-01-01"

source("run-performance.R")

p_combine / (p_ira +  p_ind) / (grow_ira + grow_ind)

```

Additionally, xxx,....xx....  dividends for the past 10 years...

```{r plot-div-hx, fig.width=10}
# instead of sourcing the function
# you can download the dev version of tidyquant
source("https://raw.githubusercontent.com/business-science/tidyquant/master/R/tq_get.R")

# plot dividends timeseries
stocks <- unique(weights$symbol)

dividends <- tq_get(stocks, 
                    get = "dividends",
                    from = start_date) %>% 
  mutate(acct = ifelse(
    symbol %in% c("EPR", "VTR"), 
    "Taxable", "Roth IRA"
  ))

dividends %>% 
  ggplot(aes(date, value)) +
  geom_line(aes(group = symbol),
            size = 1, color = "#f5ed93") +
  geom_line(data = . %>% 
              filter(symbol %in% c("VTI", "DLR")),
            aes(date, value, color = symbol)) +
  scale_x_date(breaks = scales::pretty_breaks(12)) +
  scale_y_continuous(labels = scales::dollar) +
  scale_color_manual(values = 
                       c("DLR" = "#003F4C",
                         "VTI" = "#601200")) +
  facet_wrap(~acct) +
  labs(x = NULL, color = NULL,
       y = "Dividends payout (per share)") +
  theme(panel.grid.minor = element_blank())

```


Of course, I just started this journey last month - so I'm currently seeing losses.

```{r plot-my-portfolios, fig.height=11, fig.width=12}
start_date <- "2020-04-01"

source("run-performance.R")

p_combine / (p_ira +  p_ind) / (grow_ira + grow_ind)

```

## Allocations

Here I'll assign asset class and sector category to my current holdings to get an overview of the ...... my current allocations,

```{r allocations}

allocations <- weights %>% 
  filter(acct != "combined") %>% 
  mutate(acct = ifelse(acct == "ind", 
                       "Taxable", "Roth IRA"),
         asset_class = case_when(
           symbol %in% c("VTR", "EPR", "DLR") ~ "Equity REIT",
           symbol %in% c("VTI",  "VXUS", "VNQ") ~ "Stock ETF",
           symbol == "BND" ~ "Bond ETF"
         ),
         sector = case_when(
           symbol == "VTR" ~ "Healthcare",
           symbol == "EPR" ~ "Experiential",
           symbol == "BND" ~ "US Bond market",
           symbol == "VTI" ~ "US Total market",
           symbol == "VXUS" ~ "Int'l market",
           symbol == "DLR" ~ "Data Center",
           symbol == "VNQ" ~ "Real Estate"
         ))

# donut chart ref: 
# https://www.datanovia.com/en/blog/how-to-create-a-pie-chart-in-r-using-ggplot2/
allocations %>% 
  group_by(acct) %>% 
  arrange(desc(asset_class)) %>% 
  mutate(lab.ypos = cumsum(prop) - 0.5*prop) %>%
  ggplot(aes(x = 2, y = prop, 
             fill = asset_class)) +
  geom_bar(width = 1, 
           stat = "identity", color = "white") +
  coord_polar(theta = "y", start = 0) +
  geom_text(aes(y = lab.ypos, 
                label = scales::percent(prop)), 
            color = "white") +
  ggrepel::geom_text_repel(aes(y = lab.ypos,
                               label = sector),
                           nudge_x = 1, vjust = 1, 
                           direction = "both",
                           min.segment.length = 2,
                           segment.size  = 0.2,
                           segment.colour = "Gray60") +
  gameofthrones::scale_fill_got_d() +  
  theme_void(base_family = "serif",
             base_size = 14) +
  theme(legend.position = "bottom") +
  xlim(0.5, 2.5) +
  facet_wrap(~acct)

```

we can simply rebalance a portfolio to see what it's performance could be.  

I've been really interested in holding VHT, and increasing my position in BND and VTI... You should use the actual values of your holdsings, but for simplicity, I'll multiply my current allocations by $1,000 - and add to that.

```{r rebalance}
# rebalance portfolio
rebalance <- allocations %>% 
  add_row(acct = "Taxable", symbol = "VHT", prop = 0,
          asset_class = "Stock ETF", sector = "Healthcare") %>% 
  mutate(current = prop*1000,
         new = case_when(
           symbol == "BND" ~ current + 125,
           symbol == "VTI" ~ current + 125,
           symbol == "VHT" ~ current + 200,
           TRUE ~ current + 0
         )) %>% 
  group_by(acct) %>% 
  mutate(rebalance = new / sum(new)) %>% 
  ungroup()

rebalance

```


```{r, echo=FALSE}
rebalance %>% 
  group_by(acct) %>% 
  arrange(desc(asset_class)) %>% 
  mutate(lab.ypos = cumsum(rebalance) - 0.5*rebalance) %>%
  ggplot(aes(x = 2, y = rebalance, 
             fill = asset_class)) +
  geom_bar(width = 1, 
           stat = "identity", color = "white") +
  coord_polar(theta = "y", start = 0) +
  geom_text(aes(y = lab.ypos, 
                label = scales::percent(rebalance)), 
            color = "white") +
  ggrepel::geom_text_repel(aes(y = lab.ypos,
                               label = sector),
                           nudge_x = 1, vjust = 1, 
                           direction = "both",
                           min.segment.length = 2,
                           segment.size  = 0.2,
                           segment.colour = "Gray60") +
  gameofthrones::scale_fill_got_d() +  
  theme_void(base_family = "serif",
             base_size = 14) +
  theme(legend.position = "bottom") +
  xlim(0.5, 2.5) +
  facet_wrap(~acct)

```

My returns ... blah ablaha 

```{r}
start_date <- "2020-04-01"
```

```{r, echo=FALSE}

rebalance <- allocations %>% 
  add_row(acct = "Taxable", symbol = "VHT", prop = 0,
          asset_class = "Stock ETF", sector = "Healthcare") %>% 
  mutate(prop = case_when(
           symbol == "BND" ~ .20,
           symbol == "VTI" ~ .40,
           symbol == "DLR" ~ .15,
           symbol == "VNQ" ~ .20,
           symbol == "VXUS" ~ .05,
           symbol == "EPR" ~ .50,
           symbol == "VHT" ~ .50,
         ))

new_returns <- tq_get(unique(rebalance$symbol),  
                      get = "stock.prices",
                      from = start_date) %>% 
  group_by(symbol) %>%
  tq_transmute(adjusted, periodReturn, 
               period = "monthly",
               col_rename = "Ra")

weights <- rebalance %>% select(acct, symbol, prop)

port_ira <- new_returns %>% 
  filter(!symbol %in%
           c("EPR", "VTR", "VHT")) %>% 
  get_returns("Roth IRA")

port_ind <- new_returns %>% 
  filter(symbol %in%
           c("EPR", "VTR", "VHT")) %>% 
  get_returns("Taxable")



p_ira <- new_returns %>% 
  filter(!symbol %in%
           c("EPR", "VTR", "VHT")) %>% 
  plot_returns(port_ira) +
  labs(subtitle = "Roth IRA acct.")

p_ind <- new_returns %>% 
  filter(symbol %in%
           c("EPR", "VTR", "VHT")) %>% 
  plot_returns(port_ind) +
  labs(subtitle = "Indv. taxable acct.")


(p_ira +  p_ind) 


```

